Extreme value analysis for the sample autocovariance matrices of heavy-tailed multivariate time series

### 2017-07-21 papers/DavisMikoschHeinyXie/PaperBody.tex
Added some paragraph in the introduction.

The primary objective of this overview  is to make a connection between extreme value theory and the behavior of the largest eigenvalues of sample covariance matrices from heavy-tailed multivariate time series.  For time series that are linearly dependent through time and across rows, it turns out that the extreme eigenvalues are essentially determined by the extreme order statistics from an array of iid random variables.  The asymptotic behavior of the extreme eigenvalues is then derived routinely from classical extreme value theory.  As such, explicit joint distributions of the extreme order statistics can be given which yield a plethora of ancillary results.  
%These include the limit behavior of ratios of consecutive extreme eigenvalues, ratios of extreme order statistics to the sample variance, etc..
%%The underlying multivariate \ts\ will assumed to be heavy-tailed and linearly dependent through time and across the rows.We consider asymptotic theory for the largest eigenvalues of a sample covariance matrix,
%and we will also address the largest singular values of the sample autocovariance matrix.
%To our knowledge, the present paper is the first to consider bonafide dependence among the components in the time series which renders a multivariate analysis, such as PCA, meaningful.
Convergence of the point process of extreme eigenvalues, properly normalized, plays a central role in establishing the main results.
\par
In Section~\ref{sec:2} we continue the study of the case when the data matrix $\X$ consists of iid heavy-tailed entries.  We will consider power-law growth rates on the dimension $(p_n)$ that is more general than prescribed by \eqref{eq:gamma}.
In Section~\ref{sec:model} we introduce a model for $X_{it}$ which allows for linear dependence across the rows and through time.
The point process convergence of normalized eigenvalues is presented in Section~\ref{sec:mainresult}.
This result lays the foundation for new insight into the spectral behavior of the sample covariance matrix, which is the content of Section~\ref{sec:samplecov}.

@@

\section{General growth rates for $p_n$ in the iid heavy-tailed case}\label{sec:2}\setcounter{equation}{0}
This section is based on ideas in Heiny and Mikosch \cite{heiny:mikosch:2015:iid} where one can also find detailed proofs.
\subsection*{Growth conditions on $(p_n)$}\label{subsec:pn}
In many applications it is not realistic to assume
that the dimension $p$ of the data and the sample size $n$ grow at the same rate.
The aforementioned results of Soshnikov~\cite{soshnikov:2004,soshnikov:2006} and Auffinger et al. \cite{auffinger:arous:peche:2009} already indicate that the value $\gamma$ in the growth
rate~\eqref{eq:gamma} does not appear in the \ds al limits.
This obervation is in contrast to the light-tailed case; see \eqref{eq:geman}
and \eqref{eq:tc}.  Davis et al.~\cite{davis:mikosch:pfaffel:2015,davis:pfaffel:stelzer:2014}
and Heiny and Mikosch ~\cite{heiny:mikosch:2015:iid} allowed for more general rates for
$p_n\to\infty$ than linear growth in $n$.
Recall that $p=p_n\to\infty$ is the number of rows in the
matrix $\X_n$. We need to specify the growth rate of $(p_n)$ to ensure
a non-degenerate limit distribution of the normalized singular values of the sample autocovariance
matrices. To be precise, we assume
\begin{equation}\label{eq:p}
p=p_n=n^\beta \ell(n), \quad n\ge1,\tag{$C_p(\beta)$}
\end{equation}
where $\ell$ is a slowly varying function and $\beta\ge 0$. If $\beta =0$, we also assume $\ell(n) \to \infty$.
Condition~\ref{eq:p} is more general than the growth conditions in the literature; see
\cite{auffinger:arous:peche:2009,davis:mikosch:pfaffel:2015,davis:pfaffel:stelzer:2014}.
\begin{theorem}\label{thm:intro}
Assume that $\X=\X_n$ has iid entries satisfying the \regvar\ condition \eqref{eq:regvar} for some
$\alpha \in (0,4)$. If $\E[|X|]<\infty$ we also suppose that $\E [X]=0$. Let $(p_n)$ be an integer sequence
satisfying \ref{eq:p} with $\beta\ge 0$. In addition, we require
\begin{equation}\label{Cbeta}
\min(\beta,\beta^{-1})\in (\alpha/2-1,1] \qquad \mbox{ for } \alpha\in [2,4), \tag{$\widetilde{C}_\beta(\alpha)$}
\end{equation}
Then
\beam\label{eq:2}
\sum_{i=1}^p \vep_{a_{np}^{-2}\la_i}\std \sum_{i=1}^\infty \vep_{\Gamma_i^{-2/\alpha}}\,,\qquad \nto\,,
\eeam
where the \con\ holds in the space of point \ms s with state space $(0,\infty)$ equipped with the vague toplogy; see Resnick \cite{resnick:2007}.
\end{theorem}

@@

%In  Davis et al. \cite{davis:mikosch:pfaffel;2015} relation \eqref{eq:2} was derived for $\alpha\in (0,1)$ even for \seq s
%$p_n\to\infty$ \st\ $p_n= O(\ex^{s_n/n})$, where $s_n=o(n)$ but with stronger limitations on the growth of $(p_n)$ for $[1,4]$.
%In the case $\alpha\in (2,4]$ they proved \eqref{eq:center} with $(\la_i)$ replaced by $(\la_i(\X\X'-n\E[ \X\X']))$. It is surprising that
%\eqref{eq:center} also holds since, in general, $\la_i(\X\X'-\E [\X\X'])\not = \la_i-n\,\E X$.

@@

\begin{equation}\label{eq:1}

@@

If the series in \eqref{eq:1} converges a.s. $(X_{it})$ constitutes a strictly stationary random field.

@@

the a.s.~absolute convergence of the series in \eqref{eq:1}. Under the conditions \eqref{eq:27} and \eqref{eq:2a}, the marginal and

@@

The model \eqref{eq:1} was introduced by Davis et al. \cite{davis:pfaffel:stelzer:2014}, assuming the rows iid, and in the

@@

originates from a paper by Lam and Yao~\cite{lam:yao} who used a model different from \eqref{eq:1}.

@@

the linear model \eqref{eq:1}.

@@

Consider the linear process \eqref{eq:1} under

@@

\begin{theorem}\label{cor:1}
Assume the conditions of Theorem~\ref{thm:mains}.
Then $(N_n^{\lambda,s})$ converge weakly in the space of point measures
with state space $(0,\infty)^{s+1}$ equipped with the vague topology.
If either $\alpha \in (0,2]$, $\E[Z^2]=\infty$ and $\beta \ge 0$,
or $\alpha \in [2,4)$, $\E[Z^2]<\infty$ and \ref{Cbeta} hold then
\begin{equation}\label{eq:pp}
N_n^{\lambda,s} \cid N= \sum_{i=1}^\infty
\sum_{j=1}^{\infty} \vep_{\Gamma_i^{-2/\alpha} (v_j(0),\ldots,v_j(s))}, \qquad \nto.
\end{equation}
\end{theorem}

@@

To illustrate the theory we consider a simple moving average example
taken from Davis et al.~\cite{davis:mikosch:pfaffel:2015}.
\begin{example} \rm
\begin{figure}[htb!]
  \centering
  \subfigure[iid data]{
    \includegraphics[scale=0.4]{lambda_1_2_Gaps_iid.pdf}
  }
  \subfigure[data from model \eqref{ex:4.6}]{
    \includegraphics[scale=0.4]{lambda_1_2_Gaps.pdf}
  }
  \caption{Distribution function of $(\lambda_{(1)} - \lambda_{(2)})/\lambda_{(1)}$ for iid data (left) and
    data generated from the model \eqref{ex:4.6} (right). In each graph we compare the empirical \ds\ \fct\
(dotted line, based on
1000 simulations of $200 \times 1000$ matrices with $Z$-\ds\ \eqref{eq:distrsim}) with the theoretical curve (solid line).}
  \label{fig:ProbMass}
\end{figure}
Assume that $\alpha\in (0,2)$ and
\begin{equation}\label{ex:4.6}
X_{it}= Z_{it}+ Z_{i,t-1}-2 (Z_{i-1,t}- Z_{i-1,t-1})\,,\quad i,t\in\Z\,.
\end{equation}
In this case, the non-zero entries of $\H$ are
\begin{equation*}
h_{00}=1, h_{01}=1,h_{10}=-2 \quad \mbox{ and }\quad h_{11}=2.
\end{equation*}
Hence $\M=\H\H'$ has the positive eigenvalues
$v_1=8$ and $v_2=2$. The limit point process in \eqref{eq:pp} is
\begin{equation*}
N=\sum_{i=1}^\infty \vep_{8\Gamma_i^{-2/\alpha}}+ \sum_{i=1}^\infty \vep_{2\Gamma_i^{-2/\alpha}}\,,
\end{equation*}
so that
\begin{equation*}
a_{np}^{-2}\,\big(\la_{(1)},\la_{(2)}\big) \cid
\big(8\Gamma_{1}^{-2/\alpha}, 2\Gamma_1^{-2/\alpha}\vee 8
\Gamma_2^{-2/\alpha}\big)\,.
\end{equation*}
Using the fact that $U=\Gamma_1/\Gamma_2$ has a uniform \ds\ on $(0,1)$ we calculate
\begin{equation*}
\P(2\Gamma_1^{-2/\alpha}>8\Gamma_2^{-2/\alpha})= \P(\Gamma_1/\Gamma_2<2^{-\alpha})= 2^{-\alpha} \in (1/4,1).
\end{equation*}
In particular, we have for the normalized spectral gap
\begin{equation*}
a_{np}^{-2} \big(\la_{(1)}-\la_{(2)}\big)\cid
6 \,\Gamma_1^{-2/\alpha} \1_{\{\Gamma_1 4^{\alpha/2 }<\Gamma_2\}}+
8\,
\big(\Gamma_1^{-2/\alpha}-\Gamma_2^{-2/\alpha}\big)\1_{\{\Gamma_14^{\alpha/2
  }> \Gamma_2\}}
\end{equation*}
and for the self-normalized spectral gap (see also Example~\ref{exam:spectralgap} for a detailed analysis)
\begin{equation*}
\begin{split}
\dfrac{\la_{(1)}-\la_{(2)}}{\la_{(1)}}&\cid
\dfrac{6}{8} \, \1_{\{\Gamma_1 2^{\alpha }<\Gamma_2\}}+
\big(1-(\Gamma_1/\Gamma_2)^{2/\alpha}\big)\1_{\{\Gamma_12^{\alpha
  }> \Gamma_2\}}\\
& =\dfrac{3}{4}\,  \1_{\{U 2^{\alpha }<1\}}+
\big(1-U^{2/\alpha}\big)\1_{\{U2^{\alpha
  }> 1\}}=Y\,.
\end{split}
\end{equation*}
The limit  distribution of the spectral gap
has an atom at $3/4$ with probability $2^{-\alpha}$, i.e.,~$\P(Y=3/4)=2^{-\alpha}$, and
\begin{equation*}
\P(Y\le x)= 1- (1-x)^{\alpha/2},\qquad x\in (0,3/4).
\end{equation*}
In the iid case the limit distribution of the self-normalized spectral gap has distribution function $F(x) = 1- (1-x)^{\alpha/2}$ for $x\in [0,1]$. This means that the atom disappears if the entries are iid. Figure~\ref{fig:ProbMass} compares the distribution function of $Y$
with $F$ for $\alpha=0.6$; the atom at $3/4$ is clearly visible.

@@

and hence the limit distribution of $\lambda_{(2)}/\lambda_{(1)}$ is supported on $[1/4,1)$ with  mass of $2^{-\alpha}$ at $1/4$.  The histogram of the ratio $\left(\lambda_{(2)}/\lambda_{(1)}\right)^{2/\alpha}$ based on 1000 replications from the model \eqref{ex:4.6} with noise given by a $t$-distribution with $\alpha=1.5$ degrees of freedom, $n=1000$ and $p=200$ is displayed in Figure \ref{fig:1}.

@@

\label{fig:1}

@@

%--------------------------------------------------------------------------------------
\subsection{Sums of squares of sample autocovariance matrices}\label{sec:possemidef}
In this section we consider some additive \fct s
of the squares of $\A_n(s)=\X_n(0)\X_n(s)'$ given by $\A_n(s)\A_n(s)'$ for $s=0,1,\ldots$. By definition of the singular values of a matrix
(see  \eqref{eq:sigma}), the non-negative definite
matrix $\A_n(s)\A_n(s)'$ has eigenvalues $(\la_i^2(s))_{i=1,\ldots,p}$.
\par
The following result is a corollary of Theorem~\ref{thm:mains}.
\begin{proposition}\label{thm:mainstr} Consider the linear process \eqref{eq:1} under
the conditions of Theorem~\ref{thm:mains}. Then the following statements hold for $s\ge 0$:
\begin{enumerate}
\item[$(1)$]
We consider two disjoint cases:
$\alpha \in (0,2)$ and $\beta\in (0,\infty)$, or
$\alpha\in [2,4)$ and $\beta$ satisfying \ref{Cbeta}. Then
\beao
a_{np}^{-4} \max_{i=1,\ldots,p} |\lambda_{(i)}^2(s)-\delta_{(i)}^2(s)| \cip 0, \quad \nto.
\eeao
\item[$(2)$]
Assume $\beta\in [0,1]$.
If $\alpha \in (0,2]$, $\E[Z^2]=\infty$ or $\alpha\in [2,4)$, $\E [Z^2]<\infty$ and $\beta \in (\alpha/2-1,1]$, then
\beao
a_{np}^{-4} \max_{i=1,\ldots,p} |\la_{(i)}^2(s)-(\gamma_{(i)}^\rightarrow(s))^2| \cip 0, \quad \nto.
\eeao
Assume $\beta>1$. If $\alpha \in (0,2]$, $\E[Z^2]=\infty$ or $\alpha\in [2,4)$, $\E [Z^2]<\infty$ and $\beta^{-1} \in (\alpha/2-1,1]$. Then
\beao
a_{np}^{-4} \max_{i=1,\ldots,p} |\la_{(i)}^2(s)-(\gamma_{(i)}^\downarrow(s))^2| \cip 0, \quad \nto.
\eeao
\end{enumerate}
\end{proposition}
To the best of our knowledge, sums of squares of sample autocovariance matrices were used first in the paper by Lam and Yao
\cite{lam:yao}; their \ts\ model is quite different from ours.
\begin{proof}
Part (1).
The proof follows from Theorem~\ref{thm:mains} if we can show that
\beao
a_{np}^{-2}\max_{i=1,\ldots,p} \big(\la_{(i)}(s)+\delta_{(i)}(s) \big)=O_\P(1)\,\quad \nto\,.
\eeao
We have by Theorem~\ref{cor:1},
\beam \label{eq:gdfg}
a_{np}^{-2}\max_{i=1\ldots,p} \la_{(i)}(s)=a_{np}^{-2}\la_{(1)}(s) \cid c\, \xi_{\alpha/2}\,,
\eeam
where $\xi_{\alpha/2}$ has a $\Phi_{\alpha/2}$ \ds . In view of Theorem~\ref{thm:mains}(1) we also have
\beao
a_{np}^{-2}\max_{i=1\ldots,p} \delta_{(i)}(s)\cid c\, \xi_{\alpha/2}\,.
\eeao
Therefore, again using Theorem~\ref{thm:mains}(1), we have
\beao
\lefteqn{a_{np}^{-4} \max_{i=1,\ldots,p} |\lambda_{(i)}^2(s)-\delta_{(i)}^2(s)|}\\
&\le &\big[a_{np}^{-2} \max_{i=1,\ldots,p} |\lambda_{(i)}(s)-\delta_{(i)}(s)|\big]\,
\big[a_{np}^{-2} \max_{i=1,\ldots,p}\big ( |\lambda_{(i)}(s)|+|\delta_{(i)}(s)|\big)\big]
\cip 0, \quad \nto.
\eeao
This proves part (1).\\[1mm]
Part (2). Now assume $\beta\in [0,1]$ and $\alpha \in (0,2]$, $\E[Z^2]=\infty$ or $\alpha\in [2,4)$, $\E [Z^2]<\infty$ and $\beta\in (\alpha/2-1,1]$. Then \eqref{eq:gdfg} is still true and we have by Theorem~\ref{thm:mains}(2) and Theorem~\ref{cor:1}
\beao
a_{np}^{-2}\max_{i=1\ldots,p} \gamma_{(i)}^{\rightarrow}(s)\cid c\, \xi_{\alpha/2}\,.
\eeao
We then have
\beao
\lefteqn{a_{np}^{-4} \max_{i=1,\ldots,p} |\lambda_{(i)}^2(s)-(\gamma_{(i)}^\rightarrow(s))^2|}\\
&\le &\big[a_{np}^{-2} \max_{i=1,\ldots,p} |\lambda_{(i)}(s) -\gamma_{(i)}^\rightarrow(s)|\big]\,
\big[a_{np}^{-2} \max_{i=1,\ldots,p}\big ( \lambda_{(i)}(s)+\gamma_{(i)}^\rightarrow(s)
\big)\big]
\cip 0\,, \qquad \nto.
\eeao
The proof of the remaining part is similar and therefore omitted.
\end{proof}
Now, using Proposition~\ref{thm:mainstr} and a continuous mapping argument, we can show
limit theory for the eigenvalues
\beao
w_{(1)}(s_0,s_1)\ge \cdots \ge w_{(p)}(s_0,s_1)\,,\qquad 0\le s_0\le s_1\,,
\eeao
of the non-negative definite random matrices
\begin{equation}\label{eq:sumA}
\sum_{s=s_0}^{s_1} \A_n(s) \A_n(s)'\,.
\end{equation}

>>>>

The primary objective of this overview  is to make a connection between extreme value theory and the behavior of the largest eigenvalues of sample covariance matrices from heavy-tailed multivariate time series.  For time series that are linearly dependent through time and across rows, it turns out that the extreme eigenvalues are essentially determined by the extreme order statistics from an array of iid random variables.  The asymptotic behavior of the extreme eigenvalues is then derived routinely from classical extreme value theory.  As such, explicit joint distributions of the extreme order statistics can be given which yield a plethora of ancillary results.  
%These include the limit behavior of ratios of consecutive extreme eigenvalues, ratios of extreme order statistics to the sample variance, etc..
%%The underlying multivariate \ts\ will assumed to be heavy-tailed and linearly dependent through time and across the rows.We consider asymptotic theory for the largest eigenvalues of a sample covariance matrix,
%and we will also address the largest singular values of the sample autocovariance matrix.
%To our knowledge, the present paper is the first to consider bonafide dependence among the components in the time series which renders a multivariate analysis, such as PCA, meaningful.
Convergence of the point process of extreme eigenvalues, properly normalized, plays a central role in establishing the main results.
\par
In Section~\ref{sec:Davis2} we continue the study of the case when the data matrix $\X$ consists of iid heavy-tailed entries.  We will consider power-law growth rates on the dimension $(p_n)$ that is more general than prescribed by \eqref{eq:gamma}.
In Section~\ref{sec:model} we introduce a model for $X_{it}$ which allows for linear dependence across the rows and through time.
The point process convergence of normalized eigenvalues is presented in Section~\ref{sec:mainresult}.
This result lays the foundation for new insight into the spectral behavior of the sample covariance matrix, which is the content of Section~\ref{sec:samplecov}.

@@

\section{General growth rates for $p_n$ in the iid heavy-tailed case}
\label{sec:Davis2}\setcounter{equation}{0}
This section is based on ideas in Heiny and Mikosch \cite{heiny:mikosch:2015:iid} where one can also find detailed proofs.
\subsection*{Growth conditions on $(p_n)$}\label{subsec:pn}
In many applications it is not realistic to assume
that the dimension $p$ of the data and the sample size $n$ grow at the same rate.
The aforementioned results of Soshnikov~\cite{soshnikov:2004,soshnikov:2006} and Auffinger et al. \cite{auffinger:arous:peche:2009} already indicate that the value $\gamma$ in the growth
rate~\eqref{eq:gamma} does not appear in the \ds al limits.
This obervation is in contrast to the light-tailed case; see \eqref{eq:geman}
and \eqref{eq:tc}.  Davis et al.~\cite{davis:mikosch:pfaffel:2015,davis:pfaffel:stelzer:2014}
and Heiny and Mikosch ~\cite{heiny:mikosch:2015:iid} allowed for more general rates for
$p_n\to\infty$ than linear growth in $n$.
Recall that $p=p_n\to\infty$ is the number of rows in the
matrix $\X_n$. We need to specify the growth rate of $(p_n)$ to ensure
a non-degenerate limit distribution of the normalized singular values of the sample autocovariance
matrices. To be precise, we assume
\begin{equation}\label{eq:p}
p=p_n=n^\beta \ell(n), \quad n\ge1,\tag{$C_p(\beta)$}
\end{equation}
where $\ell$ is a slowly varying function and $\beta\ge 0$. If $\beta =0$, we also assume $\ell(n) \to \infty$.
Condition~\ref{eq:p} is more general than the growth conditions in the literature; see
\cite{auffinger:arous:peche:2009,davis:mikosch:pfaffel:2015,davis:pfaffel:stelzer:2014}.
\begin{theorem}\label{thm:intro}
Assume that $\X=\X_n$ has iid entries satisfying the \regvar\ condition \eqref{eq:regvar} for some
$\alpha \in (0,4)$. If $\E[|X|]<\infty$ we also suppose that $\E [X]=0$. Let $(p_n)$ be an integer sequence
satisfying \ref{eq:p} with $\beta\ge 0$. In addition, we require
\begin{equation}\label{Cbeta}
\min(\beta,\beta^{-1})\in (\alpha/2-1,1] \qquad \mbox{ for } \alpha\in [2,4), \tag{$\widetilde{C}_\beta(\alpha)$}
\end{equation}
Then
\beam\label{eq:Davis2}
\sum_{i=1}^p \vep_{a_{np}^{-2}\la_i}\std \sum_{i=1}^\infty \vep_{\Gamma_i^{-2/\alpha}}\,,\qquad \nto\,,
\eeam
where the \con\ holds in the space of point \ms s with state space $(0,\infty)$ equipped with the vague toplogy; see Resnick \cite{resnick:2007}.
\end{theorem}

@@

%In  Davis et al. \cite{davis:mikosch:pfaffel;2015} relation \eqref{eq:Davis2} was derived for $\alpha\in (0,1)$ even for \seq s
%$p_n\to\infty$ \st\ $p_n= O(\ex^{s_n/n})$, where $s_n=o(n)$ but with stronger limitations on the growth of $(p_n)$ for $[1,4]$.
%In the case $\alpha\in (2,4]$ they proved \eqref{eq:center} with $(\la_i)$ replaced by $(\la_i(\X\X'-n\E[ \X\X']))$. It is surprising that
%\eqref{eq:center} also holds since, in general, $\la_i(\X\X'-\E [\X\X'])\not = \la_i-n\,\E X$.

@@

\begin{equation}\label{eq:Davis1}

@@

If the series in \eqref{eq:Davis1} converges a.s. $(X_{it})$ constitutes a strictly stationary random field.

@@

the a.s.~absolute convergence of the series in \eqref{eq:Davis1}. Under the conditions \eqref{eq:27} and \eqref{eq:2a}, the marginal and

@@

The model \eqref{eq:Davis1} was introduced by Davis et al. \cite{davis:pfaffel:stelzer:2014}, assuming the rows iid, and in the

@@

originates from a paper by Lam and Yao~\cite{lam:yao} who used a model different from \eqref{eq:Davis1}.

@@

the linear model \eqref{eq:Davis1}.

@@

Consider the linear process \eqref{eq:Davis1} under

@@

\begin{theorem}\label{cor:1}
Assume the conditions of Theorem~\ref{thm:mains}.
Then $(N_n^{\lambda,s})$ converge weakly in the space of point measures
with state space $(0,\infty)^{s+1}$ equipped with the vague topology.
If either $\alpha \in (0,2]$, $\E[Z^2]=\infty$ and $\beta \ge 0$,
or $\alpha \in [2,4)$, $\E[Z^2]<\infty$ and \ref{Cbeta} hold then
\begin{equation}\label{eq:Davispp}
N_n^{\lambda,s} \cid N= \sum_{i=1}^\infty
\sum_{j=1}^{\infty} \vep_{\Gamma_i^{-2/\alpha} (v_j(0),\ldots,v_j(s))}, \qquad \nto.
\end{equation}
\end{theorem}

@@

To illustrate the theory we consider a simple moving average example
taken from Davis et al.~\cite{davis:mikosch:pfaffel:2015}.
\begin{example} \rm
\begin{figure}[htb!]
  \centering
  \subfigure[iid data]{
    \includegraphics[scale=0.4]{lambda_1_2_Gaps_iid.pdf}
  }
  \subfigure[data from model \eqref{ex:4.6}]{
    \includegraphics[scale=0.4]{lambda_1_2_Gaps.pdf}
  }
  \caption{Distribution function of $(\lambda_{(1)} - \lambda_{(2)})/\lambda_{(1)}$ for iid data (left) and
    data generated from the model \eqref{ex:4.6} (right). In each graph we compare the empirical \ds\ \fct\
(dotted line, based on
1000 simulations of $200 \times 1000$ matrices with $Z$-\ds\ \eqref{eq:distrsim}) with the theoretical curve (solid line).}
  \label{fig:ProbMass}
\end{figure}
Assume that $\alpha\in (0,2)$ and
\begin{equation}\label{ex:4.6}
X_{it}= Z_{it}+ Z_{i,t-1}-2 (Z_{i-1,t}- Z_{i-1,t-1})\,,\quad i,t\in\Z\,.
\end{equation}
In this case, the non-zero entries of $\H$ are
\begin{equation*}
h_{00}=1, h_{01}=1,h_{10}=-2 \quad \mbox{ and }\quad h_{11}=2.
\end{equation*}
Hence $\M=\H\H'$ has the positive eigenvalues
$v_1=8$ and $v_2=2$. The limit point process in \eqref{eq:Davispp} is
\begin{equation*}
N=\sum_{i=1}^\infty \vep_{8\Gamma_i^{-2/\alpha}}+ \sum_{i=1}^\infty \vep_{2\Gamma_i^{-2/\alpha}}\,,
\end{equation*}
so that
\begin{equation*}
a_{np}^{-2}\,\big(\la_{(1)},\la_{(2)}\big) \cid
\big(8\Gamma_{1}^{-2/\alpha}, 2\Gamma_1^{-2/\alpha}\vee 8
\Gamma_2^{-2/\alpha}\big)\,.
\end{equation*}
Using the fact that $U=\Gamma_1/\Gamma_2$ has a uniform \ds\ on $(0,1)$ we calculate
\begin{equation*}
\P(2\Gamma_1^{-2/\alpha}>8\Gamma_2^{-2/\alpha})= \P(\Gamma_1/\Gamma_2<2^{-\alpha})= 2^{-\alpha} \in (1/4,1).
\end{equation*}
In particular, we have for the normalized spectral gap
\begin{equation*}
a_{np}^{-2} \big(\la_{(1)}-\la_{(2)}\big)\cid
6 \,\Gamma_1^{-2/\alpha} \1_{\{\Gamma_1 4^{\alpha/2 }<\Gamma_2\}}+
8\,
\big(\Gamma_1^{-2/\alpha}-\Gamma_2^{-2/\alpha}\big)\1_{\{\Gamma_14^{\alpha/2
  }> \Gamma_2\}}
\end{equation*}
and for the self-normalized spectral gap (see also Example~\ref{exam:spectralgap} for a detailed analysis)
\begin{equation*}
\begin{split}
\dfrac{\la_{(1)}-\la_{(2)}}{\la_{(1)}}&\cid
\dfrac{6}{8} \, \1_{\{\Gamma_1 2^{\alpha }<\Gamma_2\}}+
\big(1-(\Gamma_1/\Gamma_2)^{2/\alpha}\big)\1_{\{\Gamma_12^{\alpha
  }> \Gamma_2\}}\\
& =\dfrac{3}{4}\,  \1_{\{U 2^{\alpha }<1\}}+
\big(1-U^{2/\alpha}\big)\1_{\{U2^{\alpha
  }> 1\}}=Y\,.
\end{split}
\end{equation*}
The limit  distribution of the spectral gap
has an atom at $3/4$ with probability $2^{-\alpha}$, i.e.,~$\P(Y=3/4)=2^{-\alpha}$, and
\begin{equation*}
\P(Y\le x)= 1- (1-x)^{\alpha/2},\qquad x\in (0,3/4).
\end{equation*}
In the iid case the limit distribution of the self-normalized spectral gap has distribution function $F(x) = 1- (1-x)^{\alpha/2}$ for $x\in [0,1]$. This means that the atom disappears if the entries are iid. Figure~\ref{fig:ProbMass} compares the distribution function of $Y$
with $F$ for $\alpha=0.6$; the atom at $3/4$ is clearly visible.

@@

and hence the limit distribution of $\lambda_{(2)}/\lambda_{(1)}$ is supported on $[1/4,1)$ with  mass of $2^{-\alpha}$ at $1/4$.  The histogram of the ratio $\left(\lambda_{(2)}/\lambda_{(1)}\right)^{2/\alpha}$ based on 1000 replications from the model \eqref{ex:4.6} with noise given by a $t$-distribution with $\alpha=1.5$ degrees of freedom, $n=1000$ and $p=200$ is displayed in Figure \ref{fig:Davis1}.

@@

\label{fig:Davis1}

@@

%--------------------------------------------------------------------------------------
\subsection{Sums of squares of sample autocovariance matrices}\label{sec:possemidef}
In this section we consider some additive \fct s
of the squares of $\A_n(s)=\X_n(0)\X_n(s)'$ given by $\A_n(s)\A_n(s)'$ for $s=0,1,\ldots$. By definition of the singular values of a matrix
(see  \eqref{eq:sigma}), the non-negative definite
matrix $\A_n(s)\A_n(s)'$ has eigenvalues $(\la_i^2(s))_{i=1,\ldots,p}$.
\par
The following result is a corollary of Theorem~\ref{thm:mains}.
\begin{proposition}\label{thm:mainstr} Consider the linear process \eqref{eq:Davis1} under
the conditions of Theorem~\ref{thm:mains}. Then the following statements hold for $s\ge 0$:
\begin{enumerate}
\item[$(1)$]
We consider two disjoint cases:
$\alpha \in (0,2)$ and $\beta\in (0,\infty)$, or
$\alpha\in [2,4)$ and $\beta$ satisfying \ref{Cbeta}. Then
\beao
a_{np}^{-4} \max_{i=1,\ldots,p} |\lambda_{(i)}^2(s)-\delta_{(i)}^2(s)| \cip 0, \quad \nto.
\eeao
\item[$(2)$]
Assume $\beta\in [0,1]$.
If $\alpha \in (0,2]$, $\E[Z^2]=\infty$ or $\alpha\in [2,4)$, $\E [Z^2]<\infty$ and $\beta \in (\alpha/2-1,1]$, then
\beao
a_{np}^{-4} \max_{i=1,\ldots,p} |\la_{(i)}^2(s)-(\gamma_{(i)}^\rightarrow(s))^2| \cip 0, \quad \nto.
\eeao
Assume $\beta>1$. If $\alpha \in (0,2]$, $\E[Z^2]=\infty$ or $\alpha\in [2,4)$, $\E [Z^2]<\infty$ and $\beta^{-1} \in (\alpha/2-1,1]$. Then
\beao
a_{np}^{-4} \max_{i=1,\ldots,p} |\la_{(i)}^2(s)-(\gamma_{(i)}^\downarrow(s))^2| \cip 0, \quad \nto.
\eeao
\end{enumerate}
\end{proposition}
To the best of our knowledge, sums of squares of sample autocovariance matrices were used first in the paper by Lam and Yao
\cite{lam:yao}; their \ts\ model is quite different from ours.
\begin{proof}
Part (1).
The proof follows from Theorem~\ref{thm:mains} if we can show that
\beao
a_{np}^{-2}\max_{i=1,\ldots,p} \big(\la_{(i)}(s)+\delta_{(i)}(s) \big)=O_\P(1)\,\quad \nto\,.
\eeao
We have by Theorem~\ref{cor:1},
\beam \label{eq:gdfg}
a_{np}^{-2}\max_{i=1\ldots,p} \la_{(i)}(s)=a_{np}^{-2}\la_{(1)}(s) \cid c\, \xi_{\alpha/2}\,,
\eeam
where $\xi_{\alpha/2}$ has a $\Phi_{\alpha/2}$ \ds . In view of Theorem~\ref{thm:mains}(1) we also have
\beao
a_{np}^{-2}\max_{i=1\ldots,p} \delta_{(i)}(s)\cid c\, \xi_{\alpha/2}\,.
\eeao
Therefore, again using Theorem~\ref{thm:mains}(1), we have
\beao
\lefteqn{a_{np}^{-4} \max_{i=1,\ldots,p} |\lambda_{(i)}^2(s)-\delta_{(i)}^2(s)|}\\
&\le &\big[a_{np}^{-2} \max_{i=1,\ldots,p} |\lambda_{(i)}(s)-\delta_{(i)}(s)|\big]\,
\big[a_{np}^{-2} \max_{i=1,\ldots,p}\big ( |\lambda_{(i)}(s)|+|\delta_{(i)}(s)|\big)\big]
\cip 0, \quad \nto.
\eeao
This proves part (1).\\[1mm]
Part (2). Now assume $\beta\in [0,1]$ and $\alpha \in (0,2]$, $\E[Z^2]=\infty$ or $\alpha\in [2,4)$, $\E [Z^2]<\infty$ and $\beta\in (\alpha/2-1,1]$. Then \eqref{eq:gdfg} is still true and we have by Theorem~\ref{thm:mains}(2) and Theorem~\ref{cor:1}
\beao
a_{np}^{-2}\max_{i=1\ldots,p} \gamma_{(i)}^{\rightarrow}(s)\cid c\, \xi_{\alpha/2}\,.
\eeao
We then have
\beao
\lefteqn{a_{np}^{-4} \max_{i=1,\ldots,p} |\lambda_{(i)}^2(s)-(\gamma_{(i)}^\rightarrow(s))^2|}\\
&\le &\big[a_{np}^{-2} \max_{i=1,\ldots,p} |\lambda_{(i)}(s) -\gamma_{(i)}^\rightarrow(s)|\big]\,
\big[a_{np}^{-2} \max_{i=1,\ldots,p}\big ( \lambda_{(i)}(s)+\gamma_{(i)}^\rightarrow(s)
\big)\big]
\cip 0\,, \qquad \nto.
\eeao
The proof of the remaining part is similar and therefore omitted.
\end{proof}
Now, using Proposition~\ref{thm:mainstr} and a continuous mapping argument, we can show
limit theory for the eigenvalues
\beao
w_{(1)}(s_0,s_1)\ge \cdots \ge w_{(p)}(s_0,s_1)\,,\qquad 0\le s_0\le s_1\,,
\eeao
of the non-negative definite random matrices
\begin{equation}\label{eq:sumA}
\sum_{s=s_0}^{s_1} \A_n(s) \A_n(s)'\,.
\end{equation}
